---
title: "Urban Air Quality"
subtitle: "[NO_URAQ_001 to NO_URAQ_005] - Urban Air Quality - Particulate Matter 2.5 concentration in urban areas"
indicatorID: 
  - NO_URAQ_001
  - NO_URAQ_002
  - NO_URAQ_003
  - NO_URAQ_004
  - NO_URAQ_004
indicatorName: 
  - Urban Air Quality # add more by adding another line starting '-'
  verbatimName: 
  - Konsentrasjon av PM2.5 partikler
  - Concentration de particule fines PM2,5
folderName: NO_URAQ_001-005
country: Norway
continent: Europe
ECT: A2 - Chemical State characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T7 Intensive land-use biome
Ecosystem: 
  - T7.4 Urban and industrial
verbatimEcosystem: Bebygd og opparbeidet areal
yearAdded: 2025 
yearLastUpdate: 2026
status: complete 
Version: "000.002" 
VersionComment: Version 000.001 has been reviwed with new available data.
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URAQ_000-004
Normalised: No
SpatialAggregationPathway: Aggregate
Precision: No metric available #Describe the precition metric you have used. Precision may be reported either as variation among observed values (e.g. standard deviation, variance, range, interquartile range) or as uncertainty around an estimated value (e.g. standard error, confidence interval, credible interval). 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Sylvie Clappe # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz            #  Enter subsequent authors like this, or remove if not relevant
    email: balint.czucz@nina.no
    affiliations:
      - ref: myID   
AuthorList: Clappe, S., Czúcz, B.
date: July 20, 2026 # Enter date 
data_availability: silver
code_reproducibility: silver
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
bibliography: references.bib
hide: FALSE
codeReviewers:
  - name: XXXXX
    email: XXXXXX
    affiliations:
      - id: reviewerInst
        name: XXXXXX
    reviewType: code review
    github: XXXXXXX
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(terra)
library(sf)
library(dplyr)
library(readxl)
library(httr2)
library(glue)
library(tictoc)
library(duckdb)
library(duckdbfs)
library(duckspatial)
library(mirai)
library(purrr)
library(stringr)
library(exactextractr)
library(carrier)
library(DT)
library(numform)
library(conflicted)
library(giscoR)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```


```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```

```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```

::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 05 August 2025 - First draft
- 08 September 2025 - Original PR
- 08 October 2025 - First revision
- 24 October 2025 - Second revision
- 26 Nov. 2025 - Embed meta data and add open science badges
- 20 July 2026 - Reviewed code to include new available data

:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
Air quality can be measured by the presence of various pollutants in the air. Here we focused on PM2.5, i.e. the concentration of very small solid particles (particulate matter) in the air, which have a diameter under 2.5 $\mu$m. This is in line with the EU Regulation on Environmental Economic Accounting (@eu-2024-3024), which lists PM2.5 among the mandatory ecosystem condition indicators to characterise the condition of the ecosystem type *settlements and other artificial areas* (ET01, henceforward "urban ecosystems" for short). The annual concentration of PM2.5 is also a recognised national environmental indicator in Norway (Miljøindikator 4.6.2). 

In principle, there could be many other similar  air quality indicators calculated for urban ecosystems (one for each relevant pollutant). Nevertheless, these indicators would definitely be correlated with each other, and the degree to which *air quality* metrics match the definition and selection criteria for *ecosystem condition* variables is also questionable (@Czucz2025a). Accordingly, here we only present calculations for PM2.5.  

2. **Methods**:
We used three data sources: (i) the official Norwegian ecosystem map (Grunnkart) produced by NIBIO; (ii) the PM2.5 raster for the whole of Norway produced by the Norwegian Meteorological Institute, and (iii) Local Administrative Units produced by Eurostat (municipalities boundaries).

There are 2 major steps in calculating urban air quality indicators for a selected reference period and ecosystem accounting area (EAA):

1. Identify the '*target area*' for which the indicator should be calculated (cf. @Czucz2025a).

2. Calculate the mean concentration of the selected pollutant (PM2.5) within this area (for the selected accounting period).

There are several ways of defining the *target area*. In this document, we present five different variants of the indicator based on different definitions of the *target area*. Besides offering a flexible approach for performing the "right" calculation for every reporting context, these variants are also aligned across other *urban* ecosystem condition variables (URGR, TCCD), thus making it easier to establish consistency across the different variables and their policy use. 


3. **Interpretation**:
This indicator is not normalised so its values start from 0 and go above 1. The higher the value, the higher the concentration of PM2.5 in the air. Globally, we can say that a value of 0 means that the urban ecosystem is in greater ecological condition than if the value above 0. It is difficult to set meaningful ranges to determine intervals within which PM2.5 values are detrimental.

4. **Results and development needs**:
Overview of the five indicators that we calculated in this document. For a national use we recommend to use **URAQ_005**, which is most aligned with the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA; @seea-ea) among the five versions presented. Nevertheless, for official EU reporting the calculation of **URAQ_002** is the most appropriate, as long as the guidelines in the Ecosystem accounts handbook from Eurostat (@handbook2026) stay unchanged. We also provide three other indicators **URAQ_001**, **URAQ_003** and **URAQ_004**. Together with **URAQ_002**, they all use diverse ecosystem accounting areas (EAAs). Comparing them might be useful for the additional analytical insights, notably by distinguishing the different types of settlements according to their degree of urbanisation.  

* **URAQ_001**: The *target area* is the *entire area* of the municipalities considered as "cities" according to Eurostat DEGURBA classification (`DEGURBA = 1`). As the total area of Norwegian municipalities is typically dominated by natural ecosystems (forests, wetlands, mountains, etc.) this option actually contradicts key principles of ecosystem accounting (i.e., the pollutant concentrations in a forest should be considered as the condition of the forest, not that of a potentially distant urban area).

* **URAQ_002**: The *target area* is the area of the ecosystem type ET01 (*Settlements and other artificial areas*) within the municipalities considered as *"cities"* according to Eurostat's degree of urbanisation (DEGURBA) classification (`DEGURBA = 1`). This is the preferred option for calculating the mandatory Eurostat indicator according to Eurostat's guidelines (@handbook2026). 

* **URAQ_003**: The *target area* is the area of the ecosystem type ET01 (*Settlements and other artificial areas*) within the municipalities considered as "towns and suburbs" according to Eurostat's DEGURBA classification (`DEGURBA = 2`).

* **URAQ_004**: The *target area* is the area of the ecosystem type ET01 (*Settlements and other artificial areas*) within the municipalities considered as "rural areas" according to Eurostat DEGURBA classification (`DEGURBA = 3`). URAQ_003 and URAQ_004 are complementary to URAQ_002, they present an overview of the air pollution status of the three main types of municipalities in Norway. The area-weighted mean of these three indicators (URAQ_002-URAQ_004, weighted by the total area of settlements (ET01) in the respective categories) will be equal to URAQ_004 for any reporting period.  

* **URAQ_005**: The *target area* is the total area of the ecosystem type ET01 (*Settlements and other artificial areas*) in Norway. This is the option which is most aligned with the principles of SEEA EA (@seea-ea), where an ecosystem condition indicators are expected to be representative of their "parent" ecosystem type (see also @Czucz2025a). 



## 2. About the underlying data


We use the following two main datasets for our calculations: 

* **PM2.5 concentration**: Raster maps with mean annual PM2.5 concentrations at national level. The PM2.5 dataset is produced using the EMEP (large scale chemistry and transport model) and uEMEP (local dispersion model) which are validated with measurements from Norwegian air quality monitoring stations by the Norwegian Meteorological Institute (@Denby2020; @Wærsted2025). It is important to note that air quality monitoring in Norway is focused on urban areas and high-traffic locations where pollution levels are a concern.

* **Grunnkart**: Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). This map aims to implement Eurostat's EU ecosystem typology (@et-typo-eurostat) for Norway. Here we use this map for delineating the area of ecosystem type ET01 (*Settlements and other artificial areas*).

In addition we use several further administrative datasets to delineate the individual municipalities (i.e. Local Administrative Units; LAU), as well as to determine their "degree of urbanisation" (DEGURBA), and to assign them to regions (counties, "fylke").


### 2.1 Spatial and temporal resolution and extent

Data on annual mean *PM2.5 concentration* are rasters of a 100m x 100m resolution for the the whole of Norway (excluding Svalbard).

The *Grunnkart* and the *Local Administrative Units* (LAU) are vector data (polygons). Unlike traditional land cover maps (e.g. Corine Land Cover), the Grunnkart does not have a single, universal Minimum Mapping Unit (MMU). Instead, the spatial resolution of the Grunnkart is directly inherited from the underlying source datasets used in its creation. This is because the new map is a synthesis and reclassification of existing data rather than a new, independent mapping effort. As the majority of spatial information available in public datasets focuses on human settlements and infrastructure, urban ecosystems are relatively finely represented in the Grunnkart, especially compared to semi-natural and natural ecosystem types. This makes the contours of ET01 overly-detailed, often delineating individual buildings or roads passing through a forest as "urban ecosystems". This also makes the dataset somewhat unfit for its purpose (@Rendon2025) and cumbersome to process. 


### 2.2 Original units
PM2.5 concentrations are expressed in $\mu$g/m^3^, the values reflect estimated annual average concentration for each map pixel. 

### 2.3 Additional comments about the dataset
The DEGURBA dataset was used to classify the Norwegian municipalities into three categories (1: "cities", 2: "towns and suburbs", 3: "rural areas") based on their degree of urbanisation. This data are produced by Eurostat.


#### 2.3.1 Instructions for citing, using and accessing data
||Publicly accessible to download| Data and metadata available Available at|
|----|--|--------|
| **PM2.5 concentration maps**  | Yes | Description available at [met.no](^https://www.met.no/prosjekter/luftkvalitet/dokumentasjon-av-luftkvalitetsmodellen$) and in Denby et al. (2020). Data available via data catalogue at [met.no](^https://thredds.met.no/thredds/catalog/data/fou-kl/uEMEP/for_NTNU/catalog.html$)  |
**Grunnkart**  | No, only upon request to [NIBIO](^https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d$) | We used testversion 0.2, description available in @gk2 and @strand2024|
| **EU data about the municipalities**  | Yes | Data and description available from Eurostat: [LAU](^https://ec.europa.eu/eurostat/web/gisco/geodata/statistical-units/local-administrative-units$) and [DEGURBA](https://ec.europa.eu/eurostat/web/nuts/local-administrative-units) |


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)
The presence of chemical pollutants in urban ecosystems can be considered as a *pressure variable* (which then would imply that this variable should be excluded from ecosystem condition accounts). Alternatively, this variable can also be considered a *condition variable* under ECT type A2 (abiotic chemical condition characteristics).

### 3.2 Ecosystem condition characteristic
If the "local air mass" is considered to be part of the local ecosystem (as one of its "abiotic compartments"), then PM2.5 can chararcterise the chemical condition (pollutedness, intactness) of this compartment. 


### 3.3 Other standards
In this documentation, we present different variations of the air quality indicator based on PM2.5 concentration. We recommend to use **URAQ_005** as it relies upon the most ecologically correct definition, in line with the SEEA EA (@seea-ea). It however does not follow the official guidance from the European Commission for national reporting under EU Regulation 2024/3024 (@eu-2024-3024). If users need to report this indicator in an official EU reporting, we recommend to use **URAQ_002**.



### 3.4 Collinearities with other indicators
This indicator might present some correlations with the Tree Cover Density and Urban Green Areas indicators. Green urban ecosystems provide the air filtration service which allows to decrease the concentration of PM2.5 in the air. Hence a correlation between URAQ indicators and any other condition indicators directly and indirectly including green urban ecosystems in their calculations.

If additional air pollutants (e.g., tropospheric ozone) were used to quantify separate ecosystem condition variables then these variables would definitely be highly correlated.    


### 3.5 Impact factors
The concentration of PM2.5 in the air is entirely due to anthropogenic factors. A part of PM2.5 are emitted directly as solid particles from various sources, such as construction sites, unpaved roads, fields, smokestacks or fires. Another, potentially even more significant fraction of the PM2.5 particles are formed in the atmosphere from gaseous pollutants, such as sulfur dioxide or nitrogen oxides, emitted from power plants, industries, and automobiles. Vegetation, including (semi)natural ecosystem types, and urban green areas can positively impact (i.e. reduce) PM2.5 concentrations through the air filtration service provided by these ecosystems.


## 4. Reference condition and levels

### 4.1 Reference condition
In general it is difficult to set a *reference condition* for anthropogenic ecosystems. In such cases it is often recommended to consider a  theoretical "anthropogenically derived" (SEEA EA 5.72) or "best attainable" reference condition (@valecillo2022) instead, for which there is often no "prototype" available, so the best way to operationalise this type of reference condition is to determine optimal reference levels (X~100~) for each variable independently (@Czucz2025b), which is discussed below. 


### 4.2 Reference levels
Similarly to most other pollutants and other "negative" condition variables, a zero value (no pollutants present) can be considered as a meaningful value for the "optimal" *reference level* (X~100~) of PM2.5. 

In principle it could be possible to assign further reference levels (e.g. X~60~, X~0~) based on limit values (greseverdier), criteria (luftkvalitetskriterier), or national targets set by Norway's Pollution Control Regulations (Forurensningsforskriften), the Norwegian Institute of Public Health (Folkehelseinstituttet), the Norwegian Environment Agency (Miljødirektoratet), or the World Health Organization (WHO). Norway also has a national target related to the annual concentration of PM2.5, which is 8 $\mu$g/m^3^. Nevertheless, in the case of this variable, identifying the right values for the reference levels is not a purely ecologist task, and in the context of the current project the necessary interdisciplinary understanding was not achieved.  


#### 4.2.1 Spatial resolution and validity
X~100~ (the null value of PM2.5) can be seen as a universal reference level, which is valid for any occurrences and subtypes of ET01 (in any type of settlement) at any spatial resolution over all Norway. Nevertheless, if there will be any further reference levels determined for PM2.5 in the future they will also probably be universal reference levels -- nevertheless the environmental norms are defined at the local level (directly for the air quality station measurements), so if the reference levels will be linked to them, then they will also need to be applied at the local level (i.e., pixel level data will need to be rescaled first, before spatial aggregation.) 

## 5. Uncertainties
### 5.1 General uncertainties
There are several sources of uncertainties affecting the values of this indicator and its subsequent use:

- <u>data uncertainties in PM2.5</u>: these data rely on the EMEP–uEMEP modelling framework. EMEP is a regional chemistry transport model that simulates emissions, atmospheric transport, chemical transformation and deposition of pollutants, while uEMEP is a high-resolution Gaussian dispersion downscaling model that estimates local-scale contributions from emission sources such as traffic, residential wood combustion, shipping and industry. Data inputs are distributed emissions, meteorological data from a weather model, and geographic and land-surface data. Ground-truth air quality measurement from monitoring stations are used only at the validation steps of the model. Inevitably the model contains a high level of uncertainty, but these were not quantified by the data provider.

- <u>spatial, temporal and semantic uncertainties of the Grunnkart</u>: the current Grunnkart combines many different static data sources with various temporal reference dates, which does not seem to present an easily reproducible and updatable data source. This means that the outlines of the urban ecosystems (ET01) contain an unknown amount of uncertainty ("ad-hoc-ness"). The Grunnkart is currently a static dataset whose timeline has not been clearly documented as relying on temporally heterogeneous sources. Neverthless, we used it for the entire study period.

- <u>uncertainty of defining the target area</u>: the indicators presented in this document have been compiled in accordance with different interpretations of the urban areas, see more in Section 1.1. The instructions outlined in @handbook2026 do not align well with SEEA EA principles. All the URAQ indicators present a different definition of the EEA, which greatly impacts the results and is thus an important source of uncertainty.

### 5.2 Reported uncertainties
We could not quantify uncertainties linked to URAQ indicators.

## 6. References

::: {#refs}
:::


## 7. Import and prepare datasets
All the datasets used to compile the different versions of the URAQ indicators have been thoroughly described in the previous sections, links to them are also provided for more information.

We used the following five datasets: 

* **D-1**: *PM2.5 data* (met.no): a series of raster datasets used to calculate mean PM2.5 concentration over different spatial scales. 

* **D-2**: *Grunnkart* (NIBIO): a high-resolution vector dataset used to delineate urban areas (ET01: settlements and other artificial areas).

* **D-3** *LAU data* (Local Administrative Units, Eurostat GISCO): a vector dataset was used as the primary data source to delineate the  municipalities.

* **D-4** *DEGURBA*  (Degree of urbanisation) A tabular dataset assigning one of three DEGURBA classes (1: "cities"; 2: "towns and suburbs"; 3: "rural areas") to each municipality. These classes were used to breakdown the results for URAQ_001-004.

We downloaded, cleaned and formatted the data prior to compiling URAQ indicators. Section 7 only presents how to load the cleaned data.

### 7.1 PM2.5 dataset (D-1)
We have nine rasters to load, we first create a list of files before reading them and storing them in a list object. The cleaning and formatting steps only included a mask using D-3 to remove marine ecosystems. This is because the coastal municipalities have maritime territory.

```{r}
#| eval: false
#| code-summary: "Load PM2.5 data"

# Path to PM2.5
pm25_path <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/PM2.5/National"

# List the gdb files
pm25_files <- list.files(pm25_path, pattern = ".tif$", 
                         full.names = TRUE)

# Read data with gdb files
pm25_data <- pm25_files %>% 
                 map(\(.) rast(.)) 

# Name the layers
names(pm25_data) <- str_extract(pm25_files, "(?<=annualmean_).{4}")

## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pm25_data")))
```


### 7.2 Grunnkart (D-2)

### 7.3 LAU (D-3) and DEGURBA (D-4)
LAU data can be directly downloaded from Eurostat using the `giscoR` package. We download the data in temporary directory.

```{r}
#| eval: false
#| code-summary: "Load LAU data"
# Load administrative boundaries
# Create a temporary directory to store the cache
temp_dir <- tempdir()

# Download the dataset
lau_no <- gisco_get_lau(year = "2024",
                        epsg = "3035",
                        update_cache = TRUE,
                        cache_dir = temp_dir,
                        verbose = FALSE,
                        country = "NO",
                        gisco_id = NULL)
# Clean LAU codes
lau_no$GISCO_ID <- str_remove(lau_no$GISCO_ID, "NO_")

```

Now we load DEBURBA data and add it to `lau_no`.

```{r}
#| eval: false
#| code-summary: "Add DEGURBA to LAU data"

# Source function to add fylke number
source(paste0(here::here(),"/indicators/NO_URAQ_001-005/R/functions/add_fylke.R"))

# Path to DEGURBA data
path_degurba <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/Adminstrative_boundaries_NO/LAU_DEGURBA_Eurostat/EU-27-LAU-2024-NUTS-2024.xlsx"

# Read excel file with Degurba variable and select Norway
degurba_lau_no <- read_excel(path_degurba, sheet = "NO") %>%
                    mutate(Fylke_nummer = add_fylke(., kommune_nummer = "LAU CODE")[[1]],
                           Fylke_navn = add_fylke(., kommune_nummer = "LAU CODE")[[2]],)

# Join lau_no with DEGURBA
lau_dgurb <- inner_join(lau_no,
                        degurba_lau_no,
                        by = join_by("GISCO_ID" == "LAU CODE")) %>%
              select(Fylke_nummer, Fylke_navn, GISCO_ID, LAU_NAME, DEGURBA, geometry) %>%
              st_as_sf()

```

## 8. Spatial units
The indicator values of the PM2.5 variable can be interpreted at any spatial scale, including the original (100m x 100m) raster maps (basic spatial unit, BSU). The smallest policy-relevant spatial units for all variables characterising ET01 are probably the contiguous settlement areas (bydeler: ecosystem assets, EA) and municipalities. Nevertheless, coarser ecosystem accounting areas (EAA), like the *fylke* (county) or the *entire country* can also be highly relevant as reporting areas.  


## 9. Analyses


### Paths to data sources
Before starting any of the analyses, we store the path in which the data will be found. If readers wish to download the data and run this code, they can just change the paths below to their local paths for the rest of the code to function. Note: most of the data used in this script are big and required pre-processing. Some of them could be requested through API but others could not. We chose to make the code available for API requests and pre-processing here but could not make the data directly available due to their large size. 
```{r}
#| eval: true
#| code-summary: "Paths to the data"

# DEGURBA from Eurostat
path_degurba <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/Adminstrative_boundaries_NO/LAU_DEGURBA_Eurostat/EU-27-LAU-2024-NUTS-2024.xlsx"

# LAUs shapefile from Eurostat
path_lau <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/Adminstrative_boundaries_NO/LAU_Eurostat/LAU_RG_01M_2023_3035.shp"
path_lau_folder <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/Adminstrative_boundaries_NO/LAU_Eurostat/"

# Municipalities from Kartverkert
path_kom <- "/data/R/GeoSpatialData/AdministrativeUnits/Norway_AdministrativeUnits/Original/Norway_Municipalities/Administrative enheter kommuner FGDB-format/Basisdata_0000_Norge_25833_Kommuner_FGDB/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb"

# Municipalities number 2023 and 2024
path_kom_nb <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/Adminstrative_boundaries_NO/NO_Fylke/kommune_fylke_region_NO_2024.xlsx"

# Grunnkart
path_grunnkart <- "/data/R/GeoSpatialData/LandUse/Norway_Arealregnskap/Original/GrunnkartArealregnskap FGDB-format/"

# Path to functions
path_fn <- here::here("/indicators/NO_URAQ_001_005/R/functions")

# Path to write results
res_path <- paste0(get_root_NINA("P"), "412413_2023_no_egd/Results/URAQ/")

```

### Define urban areas
To be able to delineate the *target area* for the different versions of the indicator we need to create polygons delineating (i) the urban ecosystems (i.e. ET01) from all other ecosystem types, and (ii) the municipalities from each other. In addition we will need to distinguish different types of municipalities according to their "degree of urbanisation" (DEGURBA).


#### Settlements and Other Artificial Areas (ET01) within LAUs

Defining ET01 within LAUs requires the LAUs and the extent of the ET01 ecosystem type. LAUs have been created at the step before. ET01 can be retrieved from the Grunnkart produced by NIBIO (D-2). To identify these areas, we need to run a spatial intersections between the LAUs and the ET01 class from the Grunnkart. Such intersections are however time consuming so we will do it with the `duckdb` package which allows to decrease significantly the time of computation. Note that sadly, the geometry of the Grunnkart tile of Fylke 34 had to be cleaned outside of `duckdb` and using `sf` as some topology exceptions could not be corrected.

```{r}
#| eval: false
#| code-summary: "Settlements per LAUs"
# Code by Jennifer Hansen & Sylvie Clappe

## Path to data
# Grunnkart
gdb_files <- list.files(path_grunnkart, pattern = "_gdb\\.gdb$", 
                        full.names = TRUE)

# LAUs 2024
path_lau_2024 <- paste0(path_lau_folder, "LAU_EUROSTAT_2024.shp")

## Clean Grunnkart tile 34
# Clean
grunnkart_34 <- st_read(gdb_files[8], query = "SELECT * FROM arealregnskap WHERE okosystemtype_1 = 'Settlements and other artificial areas'") %>%
                  st_transform(., crs = "EPSG:3035") %>%
                  st_buffer(., 0)

# Check
which(st_is_valid(grunnkart_34) == FALSE) %>% length()


## Database setup
conn <- dbConnect(duckdb())
ddbs_install(conn)
ddbs_load(conn)


## Limit the number of threads (cores) duckdb uses
dbExecute(conn, "PRAGMA threads = 30")


## Register grunnkart_34 in the duckdb database
ddbs_write_vector(conn, grunnkart_34, "grunnkart_valid_34")



# Check that the table was registered
ddbs_list_tables(conn)

## Write SQL queries to re-project the grunnkart and make its geometry valid
query_valid <-glue("
CREATE TABLE grunnkart_valid AS
SELECT
  okosystemtype_1,
  ST_Transform(ST_MakeValid(geo), 'EPSG:25832', 'EPSG:3035', true) AS geom
  FROM grunnkart
  WHERE okosystemtype_1 = 'Settlements and other artificial areas'
")

query_valid_18_55 <- glue("
CREATE TABLE grunnkart_valid AS
SELECT
  okosystemtype_1,
  ST_Transform(ST_MakeValid(geo), 'EPSG:25833', 'EPSG:3035', true) AS geom
  FROM grunnkart
  WHERE okosystemtype_1 = 'Settlements and other artificial areas'
")

query_valid_34 <- glue("
CREATE TABLE grunnkart_valid AS
SELECT
  okosystemtype_1,
  geo AS geom
  FROM grunnkart_valid_34
  WHERE okosystemtype_1 = 'Settlements and other artificial areas'
")

query_valid_56 <- glue("
CREATE TABLE grunnkart_valid AS
SELECT
  okosystemtype_1,
  ST_Transform(ST_MakeValid(geo), 'EPSG: 25835', 'EPSG:3035', true) AS geom
  FROM grunnkart
  WHERE okosystemtype_1 = 'Settlements and other artificial areas'
")

query_intersect <- glue("
CREATE TABLE grunnkart_intersection AS
SELECT g.*, l.*
FROM grunnkart_valid g
JOIN lau_fylke l
ON ST_Intersects(g.geom, l.geom)")

query_drop <- glue("
ALTER TABLE grunnkart_intersection
DROP COLUMN geom_1")


## Intersection - source the function needed: do_intersection
source(paste0(path_fn,"/NO_URAQ_functions.R"))

# Create the input to the intersection function
numextract <- function(string){str_extract(string, "[-+]?[0-9]*\\.?[0-9]+")
}  
fylke_nb_gdb_vec <- map(gdb_files, ~ numextract(.x)) %>%
                      unlist() %>%
                      as.numeric()

# Perform the intersection
soaa_lau_inter <- map(fylke_nb_gdb_vec, 
                      ~ do_intersection(conn, .x, gdb_files, fylke_nb_gdb_vec, path_lau_2024))


# Close the connection to duckdb database (remove in-memory dB)
dbDisconnect(conn)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("path_degurba",
                    "path_kom",
                    "path_kom_nb",
                    "path_lau",
                    "path_lau_folder",
                    "path_grunnkart",
                    "res_path",
                    "soaa_lau_inter",
                    "lau_dgurb")))
```


#### Extract PM2.5 concentration in urban areas
Urban areas are defined in two ways in this project: (1) the total area of the LAU; or (2) the total area of the class "Settlements and other artificial areas" (ET01), from the EU Ecosystem Typology, within the LAUs. The mean concentration of PM2.5for case (1) is easy to get as it can be calculated based on the entire PM2.5 concentration maps that we downloaded above. For case (2) however, further manipulations are needed as we need to only extract the raster cells of the PM2.5 concentration maps that overlap with the ET01 polygons. This is what we do in the code below. Note that the function used for this (`exact_extract()`) only works if the geometry type is polygon or multipolygon. Hence, we first need to clean the ET01 maps that we created before to only extract the polygons. 

During the extraction of the raster cell values, there will be the creation of "NA" values. This means that some polygon will have a PM2.5 concentration of "NA". This is an artifact due to the raster format of the PM2.5 concentration maps. While comparing them with the polygon of the municipalities, we noticed that some municipal raster maps were smaller than their associated municipal polygon. Hence, some polygons of settlements sadly fall outside of the PM2.5 concentration map while being in the municipal polygon. This leads to the creation of "NA" values. We added a step of cleaning in the code below to remove these "NA" values.

One last note on the code below: to speed up the computation time, parallel computing has been used through the new `in_parallel()` function from the `purrr` package. This requires to make some adjustments in the code to make sure that the `workers`, or jobs running in parallel, have the objects present in  the R environment and the packages. Indeed, when using `in_parallel()`, all packages and R objects of the environment are invisible to the `workers` if not explicitly stated in the `in_parallel()` function.

```{r}
#| eval: false
#| code-summary: "Compile PM2.5 concentration in urban areas"

## Iteration variables (if the code above hasn't been run)
# List of municipalities numbers 
path_lau_2024 <- paste0(path_lau_folder, "LAU_EUROSTAT_2024.shp")
lau_2024 <- st_read(path_lau_2024)

lau_nb <- unique(lau_2024$k__2024) %>%
            sprintf("%04i",.) # 4-digits w padding zeros (e.g. "0301")

# List the years of interest
yrs <- seq(2016, 2023, 1)

## Clean ET01: keep only polygons
daemons(15)
soaa_lau <-  soaa_lau_inter %>%
              map(., in_parallel(~ sf:::st_collection_extract(.x,
                    type = "POLYGON", 
                    warn = FALSE))) %>%
              compact() %>%  
              bind_rows()

daemons(0)
  

## - source function to read the raster and extract the raster cells corresponding to SOOA : extract_rast
source(paste0(path_fn,"/NO_URAQ_functions.R"))


## Extract PM2.5 mean concentration per ET01 polygon
daemons(8)

soaa_pm25 <- map(seq_along(lau_nb), function(municipality_idx){
  
  # Filter municipality
  municipality_code <- lau_nb[municipality_idx]
  soaa_filtered <- soaa_lau %>%
                    filter(k__2024 == as.numeric(municipality_code))
  
  # Parallelise the extraction of PM2.5 cells for each year
  
  pm25_cells <- map(seq_along(yrs), in_parallel(function(j){
      extract_rast(
      path_rast_folder = "/data/P-Prosjekter2/412413_2023_no_egd/git_data/PM2.5/",
      soaa_shp = soaa_filtered,
      municipality = municipality_code,
      year = yrs[j]
    )
    }, extract_rast = extract_rast,
    soaa_filtered = soaa_filtered,
    municipality_code = municipality_code,
    yrs = yrs))
  
  return(pm25_cells)
  
  })

daemons(0)


## Clean the R environment
rm(list = setdiff(ls(), 
                  c("path_degurba",
                    "path_kom",
                    "path_kom_nb",
                    "path_lau",
                    "path_lau_folder",
                    "path_grunnkart",
                    "res_path",
                    "soaa_lau",
                    "lau_dgurb",
                    "yrs",
                    "lau_nb",
                    "soaa_pm25")))
```

#### Compile mean PM2.5 concentration in urban areas
Now that we have the PM2.5 concentration for the whole municipalities (PM2.5 raster maps), and within the ET01 of municipalities (see section above), we can calculate the mean concentration over both these areas.

```{r}
#| eval: false
#| code-summary: "Mean PM2.5 concentration in urban areas"

## Average PM2.5 in the municipality
## - source function to average PM2.5 concentration is municipality : read_and_compile_pm25
source(paste0(path_fn,"/NO_URAQ_functions.R"))

# Run the averaging function
daemons(10)

all_combinations <- expand.grid(seq_yrs = seq_along(yrs), seq_lau_nb = seq_along(lau_nb))

municip_pm25_av <- map2(all_combinations$seq_yrs, all_combinations$seq_lau_nb,
                   in_parallel(function(x, y){
                     read_and_compile_pm25(path_rast_folder = "/data/P-Prosjekter2/412413_2023_no_egd/git_data/PM2.5/",
                                           year = yrs[x], 
                                           municipality = lau_nb[y]
                                           )
                                          },
                     read_and_compile_pm25 = read_and_compile_pm25,
                     lau_nb = lau_nb, 
                     yrs = yrs
                          )
                   )
daemons(0)




##Create the raw data
# Calculate municipality total area
lau_total_area <- lau_2024 %>%
                    mutate(area_m2 = unclass(st_area(.))) %>%
                    st_drop_geometry() %>%
                    select(LAU_NAM, k__2024, DEGURBA, area_m2)

# Create raw datasets
municip_pm25_df <- municip_pm25_av %>%
                    compact() %>%
                    bind_rows() %>%
                    left_join(., lau_total_area, join_by(kommune_nummer == k__2024))


soaa_pm25_df <- soaa_pm25 %>%
                  compact() %>%
                  bind_rows() 

## Average PM2.5 in the ET01 of the municipality
soaa_pm25_av <- soaa_pm25_df %>%
                  rename(kommune_nummer = k__2024) %>%
                  select(year, LAU_NAM, kommune_nummer, DEGURBA, soaa_area_m2, mean) %>%
                  group_by(year, LAU_NAM, kommune_nummer, DEGURBA) %>%
                  summarise(mean_soaa_pm25_ug_m3 = mean(mean),
                            w_mean_soaa_pm25_ug_m3 = round2(weighted.mean(mean, soaa_area_m2), 2),
                            sd_soaa_pm25_ug_m3 = round2(sd(mean), 2),
                            count_n = n(),
                            se_soaa_pm25_ug_m3 = round2(sd(mean)/sqrt(count_n), 2),
                            total_soaa_m2 = sum(soaa_area_m2))


```

### Export result tables to derive URAQ
Due to the computation time of the code in this document, we export the tables of results. This allows us to compute and display the results in section 10. Results without having to run the whole code presented in this document.
```{r}
#| eval: false
#| code-summary: "Export result tables"

## Export raw data
write.table(soaa_pm25_df, paste0(res_path, "URAQ_SOAA_raw_data_table_10OCT2025.csv"), sep = ";", row.names = FALSE)
write.table(soaa_pm25_av, paste0(res_path, "URAQ_SOAA_summary_table_10OCT2025.csv"), sep = ";", row.names = FALSE)
write.table(municip_pm25_df, paste0(res_path, "URAQ_municipality_summary_table_10OCT2025.csv"), sep = ";", row.names = FALSE)

```


## 10. Results
<!--# Repeat the final results here. Typically this is a map or table of indicator values.
This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either. -->

From the summary table that we exported previously, we create a summary table with the following indicators:

* **URAQ_001**: Average PM2.5 concentration in the *whole area of* the municipalities considered as "cities" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 1`)

* **URAQ_002**: Average PM2.5 concentration in the urban areas (class ET01: Settlements and other artificial areas of the EU Ecosystem Typology) within the municipalities considered as "cities" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 1`)

* **URAQ_003**: Average PM2.5 concentration  in the urban areas (class ET01: Settlements and other artificial areas of the EU Ecosystem Typology) within the municipalities considered as "towns and suburbs" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 2`)
* **URAQ_004**: Average PM2.5 concentration  in the urban areas (class ET01: Settlements and other artificial areas of the EU Ecosystem Typology) within the municipalities considered as "rural areas" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 3`)

* **URAQ_005**: Average PM2.5 concentration in the urban areas of Norway (class ET01: Settlements and other artificial areas of the EU Ecosystem Typology) 

Below, we present two kinds of summary table: one national, and one presenting a breakdown by municipalities. Note that at municipality level, URAQ_001 does not present any standard error due to methodological limitations of the data.

### National Results
```{r}
#| eval: true
#| code-summary: "National summary table"

# Read exported table
soaa_res_processed <- read.table(paste0(res_path, "URAQ_SOAA_summary_table_10OCT2025.csv"), sep = ";", head = TRUE)
municip_res_processed <- read.table(paste0(res_path, "URAQ_municipality_summary_table_10OCT2025.csv"), sep = ";", head = TRUE)

# Compile the different URAQ indicators at national level
uraq_001_nat <- municip_res_processed %>%
                select(year, kommune_nummer, DEGURBA, mean_pm25_municip_ug_m3, area_m2) %>%
                filter(DEGURBA == 1) %>%
                group_by(year) %>%
                summarise(w_mean_pm25_nat = round2(weighted.mean(mean_pm25_municip_ug_m3, area_m2), 2),
                          count_n = n(), 
                          se_pm25_nat = round2(sd(mean_pm25_municip_ug_m3)/sqrt(count_n), 2)) %>%
                mutate(ID = "URAQ 001") 

uraq_002_nat <- soaa_res_processed %>%
                select(year, DEGURBA, mean_soaa_pm25_ug_m3, total_soaa_m2) %>%
                filter(DEGURBA == 1) %>%
                group_by(year) %>%
                summarise(w_mean_pm25_nat = round2(weighted.mean(mean_soaa_pm25_ug_m3, total_soaa_m2), 2),
                          count_n = n(),
                          se_pm25_nat = round2(sd(mean_soaa_pm25_ug_m3)/sqrt(count_n), 2)) %>%
                mutate(ID = "URAQ 002")

uraq_003_nat <- soaa_res_processed %>%
                select(year, DEGURBA, mean_soaa_pm25_ug_m3, total_soaa_m2) %>%
                filter(DEGURBA == 2) %>%
                group_by(year) %>%
                summarise(w_mean_pm25_nat = round2(weighted.mean(mean_soaa_pm25_ug_m3, total_soaa_m2), 2),
                          count_n = n(),
                          se_pm25_nat = round2(sd(mean_soaa_pm25_ug_m3)/sqrt(count_n), 2)) %>%
                mutate(ID = "URAQ 003")

uraq_004_nat <- soaa_res_processed %>%
                select(year, DEGURBA, mean_soaa_pm25_ug_m3, total_soaa_m2) %>%
                filter(DEGURBA == 3) %>%
                group_by(year) %>%
                summarise(w_mean_pm25_nat = round2(weighted.mean(mean_soaa_pm25_ug_m3, total_soaa_m2), 2),
                          count_n = n(),
                          se_pm25_nat = round2(sd(mean_soaa_pm25_ug_m3)/sqrt(count_n), 2)) %>%
                mutate(ID = "URAQ 004")

uraq_005_nat <- soaa_res_processed %>%
                select(year, mean_soaa_pm25_ug_m3, total_soaa_m2) %>%
                group_by(year) %>%
                summarise(w_mean_pm25_nat = round2(weighted.mean(mean_soaa_pm25_ug_m3, total_soaa_m2), 2),
                          count_n = n(),
                          se_pm25_nat = round2(sd(mean_soaa_pm25_ug_m3)/sqrt(count_n), 2)) %>%
                mutate(ID = "URAQ 005")

res_national <- bind_rows(uraq_001_nat,
                          uraq_002_nat,
                          uraq_003_nat,
                          uraq_004_nat,
                          uraq_005_nat) %>%
                mutate(Indicator = "Urban Air Quality",
                       Accounting_area = "Norway") %>%
                select(Indicator, ID, Accounting_area, year, w_mean_pm25_nat, se_pm25_nat)

colnames(res_national) <- c("Indicator",
                            "ID",
                            "Accounting area",
                            "Accounting period",
                            "Average PM2.5 concentration µg/m3",
                            "Standard error (µg/m3)")

datatable(res_national, filter = "top")
```

### Municipality level
Note that URAQ_004 is not reported here as  this indicator is only for national level. An area-weighted mean of URAQ_001 to URAQ_003 would sum to URAQ_004.
```{r}
#| eval: true
#| message : false
#| code-summary: "Municipal summary table"

# Read exported table
soaa_res_processed <- read.table(paste0(res_path, "URAQ_SOAA_summary_table_10OCT2025.csv"), sep = ";", head = TRUE)
municip_res_processed <- read.table(paste0(res_path, "URAQ_municipality_summary_table_10OCT2025.csv"), sep = ";", head = TRUE)

# Compile the different URAQ indicators at regional level
uraq_001_mun <- municip_res_processed %>%
                select(year, kommune_nummer, LAU_NAM, DEGURBA, mean_pm25_municip_ug_m3) %>%
                mutate(w_mean_pm25_ug_m3 = round2(mean_pm25_municip_ug_m3, 2),
                       se_pm25_ug_m3 = NA,
                       ID = "URAQ 001") %>%
                select(!mean_pm25_municip_ug_m3)

uraq_002_mun <- soaa_res_processed %>%
                select(year, kommune_nummer, LAU_NAM, DEGURBA, w_mean_soaa_pm25_ug_m3, se_soaa_pm25_ug_m3) %>%
                rename(w_mean_pm25_ug_m3 = w_mean_soaa_pm25_ug_m3,
                        se_pm25_ug_m3 = se_soaa_pm25_ug_m3) %>%
                filter(DEGURBA == 1) %>%
                mutate(ID = "URAQ 002")

uraq_003_mun <- soaa_res_processed %>%
                select(year, kommune_nummer, LAU_NAM, DEGURBA, w_mean_soaa_pm25_ug_m3, se_soaa_pm25_ug_m3) %>%
                rename(w_mean_pm25_ug_m3 = w_mean_soaa_pm25_ug_m3,
                        se_pm25_ug_m3 = se_soaa_pm25_ug_m3) %>%
                filter(DEGURBA == 2) %>%
                mutate(ID = "URAQ 003")

uraq_004_mun <- soaa_res_processed %>%
                select(year, kommune_nummer, LAU_NAM, DEGURBA, w_mean_soaa_pm25_ug_m3, se_soaa_pm25_ug_m3) %>%
                rename(w_mean_pm25_ug_m3 = w_mean_soaa_pm25_ug_m3,
                        se_pm25_ug_m3 = se_soaa_pm25_ug_m3) %>%
                filter(DEGURBA == 3) %>%
                mutate(ID = "URAQ 004")

res_municip <- bind_rows(uraq_001_mun,
                         uraq_002_mun,
                         uraq_003_mun,
                         uraq_004_mun) %>%
              mutate(Indicator = "Urban Air Quality") %>%
              select(Indicator, ID, kommune_nummer, LAU_NAM, year, w_mean_pm25_ug_m3, se_pm25_ug_m3)

res_municip$se_pm25_ug_m3 <- as.character(res_municip$se_pm25_ug_m3) %>%
                                replace_na("Not available")

colnames(res_municip) <- c("Indicator",
                           "ID",
                           "Accounting area (municipality number)",
                           "Accounting area (municipality name)",
                           "Accounting period",
                           "Average PM2.5 concentration (µg/m3)",
                           "Standard error (µg/m3)")


datatable(res_municip, filter = "top")
```



## 11. Export file


Finally we present a sample code which can export the results into csv or xlsx files: 

```{r export}
#| eval: false
#| code-summary: "Code to export the result tables above"

#CSV
write.table(res_national, "URAQ_NO_national_2016_2023.csv")
write.table(res_municip, "URAQ_NO_regional_2016_2023.csv")

# #Excel
# write_xlsx(res_national, "URAQ_NO_national_2016_2023.xlsx")
# write_xlsx(res_municip, "URAQ_NO_regional_2016_2023.xlsx")

```


::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::